---
title: "Project 1"
author: "Javi Brito"
format: 
  html:
    theme: solar
editor: visual
---

```{r libraries, echo = FALSE, include = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
library(tibble)
library(lubridate)
library(hms) #supports construction from time values
```

## Introduction

This project focuses on building R functions which query the U.S Census Berenau's Public Use Microdata Sample (PUMS) API directly. Our objective is to practice working with REST API content by constructing request URLs, parsing raw JSON responses, and converting the results into useful type datatype.

## Variables

## Variables Used

This project works with a subset of the American Community Survey PUMS variables, listed below with a short description of what each one represents.

**Numeric variables**

| Variable | Description |
|--------------------------------|----------------------------------------|
| `AGEP` | Age, in years |
| `GASP` | Monthly gas cost for the household |
| `GRPIP` | Gross rent as a percentage of household income (past 12 months) |
| `JWAP` | Time of arrival at work, returned as a coded time range and converted here to a single time-of-day value |
| `JWDP` | Time of departure for work, same coded-range treatment as `JWAP` |
| `JWMNP` | Travel time to work, in minutes |
| `PWGTP` | Person weight: the number of people in the population this row represents. Always included automatically, since every weighted statistic in this project depends on it |

**Categorical variables**

| Variable | Description                                             |
|----------|---------------------------------------------------------|
| `FER`    | Whether the person gave birth within the past 12 months |
| `HHL`    | Household language: the primary language spoken at home |
| `SCH`    | School enrollment status                                |
| `SCHL`   | Educational attainment                                  |
| `SEX`    | Sex                                                     |

**Geography**

| Level      | Description                                                 |
|------------|-------------------------------------------------------------|
| `Region`   | One of four Census regions: Northeast, Midwest, South, West |
| `Division` | A finer breakdown than Region (nine Census divisions total) |
| `State`    | Individual U.S. state, identified by FIPS code              |

## Part 1: Data Processing

### Parsing Variable Values

```{r extract-item-labels}
# Census variables.json nests value labels under a list named "item"
# this helper unwraps that structure into a simple code aka label list.
extractItemLabels <- function(values) {
  if (is.null(values)) {
    return(list())
  }

  if (is.list(values) && !is.null(values$item)) {
    values <- values$item
  }

  if (is.list(values) && length(values) == 1L && is.list(values[[1]])) {
    values <- values[[1]]
  }

  values
}
```

```{r get-variable-values}
getVariableValues <- function(year) {
  if (!year %in% 2021:2024) {
    stop("Year must be 2021, 2022, 2023, or 2024")
  }

  # the state geography variable is named "ST" in 2021/2022 and "STATE" in 2023/2024 
  geo_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"

  # only gets dictionary entries for variables this project actually needs
  vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                    geo_var, "REGION", "DIVISION")

  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json")
  resp <- GET(url)

  # simplifyVector = FALSE keeps the JSON's nested list structure intact,
  # which extractItemLabels() needs to detect "item" wrappers below
  parsed <- fromJSON(rawToChar(resp$content), simplifyVector = FALSE)
  all_vars <- parsed$variables

  # build the label mapping for each variable we need
  value_list <- list()
  for (v in vars_needed) {
    value_list[[v]] <- extractItemLabels(all_vars[[v]]$values)
  }

  value_list
}
```

```{r cache-variable-values}
# Build the variable dictionary once per year and cache it to disk, so we
# don't hit the Census API every time this document is knit
if (!file.exists("data/variable_values.rds")) {
  years <- c(2021, 2022, 2023, 2024)
  variable_values <- list()
  for (yr in years) {
    variable_values[[as.character(yr)]] <- getVariableValues(yr)
  }
  dir.create("data", showWarnings = FALSE)  # showWarnings = FALSE: fine if "data/" already exists
  saveRDS(variable_values, "data/variable_values.rds")
}
```

```{r load-variable-values}
# Loads the cached dictionary for one year a list of code = label
# mappings, one per variable for use by getSingleYearData()
loadVariableValues <- function(year) {
  variable_values <- readRDS("data/variable_values.rds")
  variable_values[[as.character(year)]]
}
```

```{r sanity-check-SCH}
# quick sanity check that the dictionary loaded and looks right
loadVariableValues(2021)$SCH
```

### Parsing API Data

Firstly, we wanted to confirm we could get the data back from the Census API with a single hardcoded request given in our assignment. This meant working out the request format directly by adding the parameters in the URL along with the attached key, while expecting a successful response, before trying to generalize any of it into a function.

We stored the API key as an environment variable. The test URL above filters to North Carolina (state:37) and adds SCHL = 24 (a doctorate student) in order to keep the return rows small and easy to verify without a specific format yet. We ran a sanity check with status_code() returning a 200 meaning that the request went through. We still do not know if the data is usable, but that is future problem.

```{r test-basic-request}
census_key <- Sys.getenv("CENSUS_KEY")
if (!nzchar(census_key)) {
  stop("CENSUS_KEY was not found. Add it to your .Renviron file and restart R.")
}

# given URL state = 37 (NC)
census_url <- paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  "?get=AGEP,PWGTP,SEX",
  "&for=state:37",
  "&SCHL=24",
  "&key=", census_key
)

test_response <- GET(census_url)
status_code(test_response)
```

That previous return needed its own function fromJSON(), which parses the API response into a plain R matrix (not even a dataframe). We used the response_to_tibble() to convert a matrix in an easy readable tibble, which shows the first row to actual column names and drops it, returning what as users consider tabular data. Since the request is working, the next step was to build getSingleYearData(), starting with validating each argument before it is ever used to build a request.

```{r response-to-tibble}
response_to_tibble <- function(response) {
  # stop if the API request was unsuccessful
  stop_for_status(response)

  # parse the JSON response
  api_data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

  # api_data returns as a matrix, we convert it in a tibble first,
  # then promote its first row to the real column names and drop that row
  api_data <- as_tibble(api_data, .name_repair = "unique")

  api_data |>
    setNames(unlist(slice(api_data, 1), use.names = FALSE)) |>
    slice(-1)
}
```

The year argument must only accept a single value from 2021 through 2024. Anything else fails immediately rather sending a request the API would reject regardless.

```{r check-year}
# year of survey: single value, defaults to 2024, must be 2021-2024
year <- 2024

if (length(year) != 1 || !year %in% 2021:2024) {
  stop("year must be a single value: 2021, 2022, 2023, or 2024")
}
```

The numeric variables needed two layers of validation. First, the requested names have to come from the allow list, and PWGTP is forced to be included even if the user does not ask for it, since none of the summary or plotting functions would work properly without it. Also, it requires at least one other numeric variable, because a request that only returned weights would not be useful on its own.

```{r check-numeric-vars}
numeric_vars <- c("AGEP", "PWGTP")  # default

# only these six variables are valid choices
# PWGTP is handled separately below as it is always included, not optional
allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")

if (!all(numeric_vars %in% c(allowed_numeric, "PWGTP"))) {
  stop("numeric_vars must be from: ", paste(allowed_numeric, collapse = ", "))
}

# PWGTP must always be returned, whether or not the user asked for it
if (!("PWGTP" %in% numeric_vars)) {
  numeric_vars <- c(numeric_vars, "PWGTP")
}

# at least one numeric variable besides PWGTP is required
if (length(numeric_vars) < 2) {
  stop("you must request at least one numeric variable besides PWGTP")
}
```

There are two numeric variables, `JWAP` and `JWDP` that do not come back directly as numbers, instead they returned as coded time-range variables like "12:00pm to 12:05pm". In order to convert this into a simple single numeric value, we took the midpoint of each range and returned it as a time-of-day value using the `library(hms)`. For those labels that show that the person does not have an applicable arrival/departure time come back as NA instead of a formatted time.

```{r get-time-midpoint}

# JWAP/JWDP come back as coded time-range labels ("12:00 p.m. to 12:04 p.m.") instead of plain numbers 
# this function finds  the midpoint of that range as a time value
# and returns NA for missing/"N/A" labels

getTimeMidpoint <- function(label) {
  if (is.na(label) || grepl("N/A", label)) return(NA)
  label <- gsub("\\.", "", label)
  times <- strsplit(label, " to ")[[1]]
  start_time <- parse_date_time(times[1], orders = "I:M p")
  end_time <- parse_date_time(times[2], orders = "I:M p")
  if (end_time < start_time) end_time <- end_time + days(1)
  midpoint <- start_time + (end_time - start_time) / 2
  as_hms(midpoint)
}

getTimeMidpoint("12:00 p.m. to 12:04 p.m.")
getTimeMidpoint("N/A (not a worker; worker who worked from home)")
```

The function covertNumericVar() decides per each column, whether a variable needs a time-midpoint calculation or just needs to be forced to a plain number. For example, with the API's "N" not-applicable code converted to a NA first.

```{r convert-numeric-var}
# applies the right conversion to one requested numeric column: JWAP/JWDP
# get the time-midpoint treatment above 
# every other numeric variable just becomes a plain number ("N" -> NA)
convertNumericVar <- function(values, var_name, var_values) {
  if (var_name %in% c("JWAP", "JWDP")) {
    labels <- var_values[[var_name]]
    midpoints <- rep(NA, length(values))
    for (i in seq_along(values)) {
      label <- labels[[values[i]]]
      if (!is.null(label)) midpoints[i] <- as.numeric(getTimeMidpoint(label))
    }
    return(as_hms(midpoints))
  }
  values[values == "N"] <- NA
  as.numeric(values)
}
```

We also needed to check for categorical variables go through the same two part validation, given that we needed to restrict to an allowed list, with at least one required SEX is just a sensible default. Unlike PWFTP, there is no categorical variable that gets force-included.

```{r check-categorical-vars}
# specify categorical variables to return
# SEX is the default
categorical_vars <- c("SEX")
allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")

# error on any variable name outside the allowed list
if (!all(categorical_vars %in% allowed_categorical)) {
  stop("categorical_vars must be from: ", paste(allowed_categorical, collapse = ", "))
}

# at least one categorical variable is required
if (length(categorical_vars) < 1) {
  stop("you must request at least one categorical variable")
}
```

Turning a raw categorical column into something useful is useful for this project so we can convert into a factor with real labels rather than leaving it as bare integer codes. convertCategoricalVar() uses the dictionary we built in the "Parsing Variable Values" section.

```{r convert-categorical-var}
# Turns a raw categorical column into a factor, using the code -> label
# lookup pulled from that year's variable dictionary (loadVariableValues())
convertCategoricalVar <- function(values, var_name, var_values) {
  labels <- var_values[[var_name]]
  factor(values, levels = names(labels), labels = unlist(labels))
}
```

We are also given that geography level is restricted to exactly one of Region, Division, or State. All three is not a valid level, if the user asks for "all" geographies only makes sense once we have already picked a level.

```{r check-geo-level}
# level of geography to pull data for
# State is the default
geo_level <- "State"
allowed_geo_levels <- c("Region", "Division", "State")

# exactly one value, must be one of the three levels above
# Note: "All" is NOT valid here (that's only allowed for geo_values, below)
if (length(geo_level) != 1 || !geo_level %in% allowed_geo_levels) {
  stop("geo_level must be exactly one of: ", paste(allowed_geo_levels, collapse = ", "))
}
```

Here is when I wanted to throw my laptop from the window. There is this ST/STATE name inconsistency, this time for validating the actual geography codes rather than just building the URL. The dictionary key for state-level data is ST in 2021 and 2022 and STATE from 2023 onward. Unlike geo_level and geo_values, we can allow all as a special case, which ignores the dictionary validation completely rather than trying to match it against all the codes.

```{r check-geo-value}
# default geography value. example default is a single State
geo_values <- "37"  # North Carolina

if (!identical(geo_values, "all")) {
  # dictionary field name for this geography
  # STATE used "ST" in the 2021/2022 dictionaries and "STATE" from 2023 on
  # REGION/DIVISION didn't change name across years
    
  geo_field <- if (geo_level == "State" && year %in% c(2021, 2022)) {
    "ST"
  } else {
    toupper(geo_level)
  }

  var_values <- loadVariableValues(year)
  valid_codes <- names(var_values[[geo_field]])

  if (!all(geo_values %in% valid_codes)) {
    stop("geo_values must be \"all\" or one of: ", paste(valid_codes, collapse = ", "))
  }
}
```

After every piece validated individually above, we created `getSingleYearData()` which assembles all of it into one big function. Build the request URL from the validated inputs, send it parse the response into a tibble, convert each column to its proper type, and tag the result with a "census" class so the custom summary()/plot() methods we built in Part 2 know how to handle it.

```{r get-single-year-data}
getSingleYearData <- function(year = 2024,
                               numeric_vars = c("AGEP"),
                               categorical_vars = c("SEX"),
                               geo_level = "State",
                               geo_values = "37") {

  # year of survey 
  if (length(year) != 1 || !year %in% 2021:2024) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024")
  }

  # numeric variables 
  allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if (!all(numeric_vars %in% allowed_numeric)) {
    stop("numeric_vars must be from: ", paste(allowed_numeric, collapse = ", "))
  }
  if (!("PWGTP" %in% numeric_vars)) {
    numeric_vars <- c(numeric_vars, "PWGTP")
  }
  if (length(numeric_vars) < 2) {
    stop("you must request at least one numeric variable besides PWGTP")
  }

  # categorical variables 
  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (!all(categorical_vars %in% allowed_categorical)) {
    stop("categorical_vars must be from: ", paste(allowed_categorical, collapse = ", "))
  }
  if (length(categorical_vars) < 1) {
    stop("you must request at least one categorical variable")
  }

  # geography level 
  allowed_geo_levels <- c("Region", "Division", "State")
  if (length(geo_level) != 1 || !geo_level %in% allowed_geo_levels) {
    stop("geo_level must be exactly one of: ", paste(allowed_geo_levels, collapse = ", "))
  }

  # lowercase keyword the API's for= expects, regardless of year
  geo_url_field <- tolower(geo_level)

  # dictionary/column field name:STATE used "ST" in the 2021/2022
  # dictionaries and "STATE" from 2023 on
  # REGION/DIVISION kept one name
  geo_field <- if (geo_level == "State" && year %in% c(2021, 2022)) {
    "ST"
  } else {
    toupper(geo_level)
  }

  # geography value(s) 
  if (!identical(geo_values, "all")) {
    var_values <- loadVariableValues(year)
    valid_codes <- names(var_values[[geo_field]])
    if (!all(geo_values %in% valid_codes)) {
      stop("geo_values must be \"all\" or one of: ", paste(valid_codes, collapse = ", "))
    }
  }

  # build the request and pull the data 
  all_vars <- c(numeric_vars, categorical_vars)
  geo_value_str <- paste(geo_values, collapse = ",")

  census_url <- paste0(
    "https://api.census.gov/data/", year, "/acs/acs1/pums",
    "?get=", paste(all_vars, collapse = ","),
    "&for=", geo_url_field, ":", geo_value_str,
    "&key=", Sys.getenv("CENSUS_KEY")
  )

  response <- GET(census_url)
  my_tibble <- response_to_tibble(response)

  # convert variable types 
  var_values <- loadVariableValues(year)

  for (v in numeric_vars) {
    my_tibble[[v]] <- convertNumericVar(my_tibble[[v]], v, var_values)
  }
  for (v in categorical_vars) {
    my_tibble[[v]] <- convertCategoricalVar(my_tibble[[v]], v, var_values)
  }

  # tag the class and return 
  class(my_tibble) <- c("census", class(my_tibble))
  my_tibble
}
```

To verify the assembled function actually works end to end we ran it against a real request and inspected both the class and the resulting data directly.

```{r test-valid-call}
# a normal call should return a census-classed tibble
test_data <- getSingleYearData(year = 2024, numeric_vars = "AGEP",
                                categorical_vars = "SEX",
                                geo_level = "State", geo_values = "37")
class(test_data)
test_data
```

Wrapping each bad-input case in tryCatch() lets the document show the exact error message produced without the whole render stopping partway through such as an invalid year, an invalid geography level, and a geography value that does not exist in that year's dictionary all get caught here.

```{r test-error-handling}
# invalid year
tryCatch(
  getSingleYearData(year = 2020),
  error = function(e) cat("Caught expected error:", conditionMessage(e), "\n")
)

# invalid geo_level ("All" is not allowed as a level, only as a value)
tryCatch(
  getSingleYearData(geo_level = "All", geo_values = "all"),
  error = function(e) cat("Caught expected error:", conditionMessage(e), "\n")
)

# geo_values not in the dictionary for that year
tryCatch(
  getSingleYearData(geo_values = "999"),
  error = function(e) cat("Caught expected error:", conditionMessage(e), "\n")
)
```

### Combining Years

Our function getSingleYearData() is only able to return one year at the time, which is not useful when we want to investigate changes overtime. Instead of duplicating any of the request/parsing/conversion logic, we built combineYearData() which calls getSingleData() one per request year and stacks the results together.

This function has a couple of details worth to highlight. Each individual year's tibble returns already tagged with the "census" class, but we strip that off before combining them. The bind_rows() does not need to know about a custom class, and re-tagging happens once, on the final combined result, rather than on each piece. We also attach a year column to each year's data before stacking since once everything is combined there's no way to know which rows came from which year. Finally, I convert year into a factor with its levels explicitly sorted, so that later when this combined data gets passed to plot() or summary(), years come out in chronological order rather than however bind_rows() stacked them.

```{r combine-years-data}
combineYearsData <- function(years,
                              numeric_vars = c("AGEP"),
                              categorical_vars = c("SEX"),
                              geo_level = "State",
                              geo_values = "37") {

  # checks years the same way getSingleYearData() validates a single
  # year, but allow more than one value here
  if (length(years) < 1 || any(!years %in% 2021:2024)) {
    stop("years must contain one or more of 2021, 2022, 2023, or 2024")
  }
  years <- unique(years)

  # pull one year at a time by reusing getSingleYearData() 
  # takes the "census" class before combining so bind_rows() treats each piece as plain tibble
  # and tag on a year column so we don't losec count on the rows came from which year once everything # is stacked together
  all_years <- lapply(years, function(yr) {
    one_year <- getSingleYearData(
      year = yr,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geo_level = geo_level,
      geo_values = geo_values
    )
    class(one_year) <- setdiff(class(one_year), "census")
    one_year$year <- yr
    one_year
  })

  # saves all years into a single tibble, make year an ordered factor so
  # it displays/plots in chronological 
  # then re-tag the class now that the combined data is ready to use
  combined <- bind_rows(all_years)
  combined$year <- factor(combined$year, levels = as.character(sort(years)))
  class(combined) <- c("census", class(combined))
  combined
}
```

In testing this function the default arguments for `combineYearsData()` were written as `numeric_vars = c("AGEP", "PWGTP")` and `geo_level = "state"` (lowercase), both of which immediately fail when passed into `getSingleYearData()`. PWGTP is not a valid explicit choice for numeric_vars, as it is automatically added, and geo_level validation is case-sensitive.

```{r test-combine-valid}
combined_test <- combineYearsData(years = c(2024, 2023))
class(combined_test)
combined_test
```

In addition to the two tryCatch() cases below, we wanted to check the quieter behavior of the function. Passing the same year multiple times shouldn't throw an error at all, but should just silently collapse to one year's worth of data. This final check here confirms that by comparing row counts instead of expecting an error message.

```{r test-combine-error-handling}
# invalid year in the vector
tryCatch(
  combineYearsData(years = 2020),
  error = function(e) cat("Caught expected error:", conditionMessage(e), "\n")
)

# empty years vector
tryCatch(
  combineYearsData(years = integer(0)),
  error = function(e) cat("Caught expected error:", conditionMessage(e), "\n")
)

# duplicate years should be silently de-duplicated
# confirm the row count matches one year's worth of data, not three
dup_test <- combineYearsData(years = c(2024, 2024, 2024))
nrow(dup_test)
```

## Part 2: Write a Generic Function for Summarizing Data

### Generic Functions

Before we start writing our customs summary() and plot() behavior from this data, it is good to understand how R works behind the screen and decides which version of a function to actually run. Given both `summary()` and `plot()` are genetic functions in R, meaning that they do not work by themselves, they need to look at the class of the object was passed in and dispatch to a class-specific method.

```{r explore-generic-dispatch}
plot.function                        # method used when class = "function"
getS3method("plot", "data.frame")    # method used when class = "data.frame"
```

The generic function plot() calls UseMethod("plot") and then R chooses plot.function or plot.data.frame depending on the class of the argument. Since the tibbles returned by getSingleYearData() and combineYearData() are tagged to the "census" class, these defined the function named summary.census and plot.census, then if we use summary(my_data) and plot(my_data) to use them instead of just their default versions.

### Numeric Summaries

The rows in this data do not each represent one person. The variable PWGT explains how many people a given row actually stands for it, meaning it is a plain mean or count across rows would misrepresent the real population. The summary.census function is built so every statistic it produces is weighted by PWGTP instead of computed on raw values.

```{r summary-census}
summary.census <- function(object, numeric_vars = NULL, categorical_vars = NULL, ...) {

  # weighted summaries require the person-weight column
  if (!"PWGTP" %in% names(object)) {
    stop("object must have a PWGTP column to compute weighted summaries")
  }
  weight_vector <- object$PWGTP

  # default every numeric column except PWGTP itself (this also naturally
  # skips JWAP/JWDP, since those are stored as hms/time values, not numeric)
  if (is.null(numeric_vars)) {
    is_num <- vapply(object, is.numeric, logical(1))
    numeric_vars <- setdiff(names(object)[is_num], c("PWGTP", "JWAP", "JWDP"))
  }

  # default  every factor column, which naturally includes "year" when present
  if (is.null(categorical_vars)) {
    is_cat <- vapply(object, is.factor, logical(1))
    categorical_vars <- names(object)[is_cat]
  }

  # weighted mean/sd for each numeric variable
  # Var(X) = E[X^2] - E[X]^2, with weights standing in for the sampling
  # probabilities each row represents
  numeric_summary <- list()
  for (var in numeric_vars) {
    numeric_vector <- object[[var]]
    sample_mean <- sum(numeric_vector * weight_vector) / sum(weight_vector)
    sample_sd <- sqrt(sum(numeric_vector^2 * weight_vector) / sum(weight_vector) - sample_mean^2)
    numeric_summary[[var]] <- list(mean = sample_mean, sd = sample_sd)
  }

  # weighted counts (estimated population totals, not raw row counts)
  # for each categorical variable
  categorical_summary <- list()
  for (var in categorical_vars) {
    categorical_summary[[var]] <- tapply(weight_vector, object[[var]], sum)
  }

  list(numeric = numeric_summary, categorical = categorical_summary)
}
```

### Graphical Summaries

This same idea of weighting carries over to plotting by the plot.census function which requires the user to specify one categorical and one numeric variable. Then, it produces a boxplot of the numeric variable across the levels of the categorical one. We set weight = PWGTP inside in the aes() so the boxplot reflects the real weighted population distribution rather than the sampled rows.

```{r plot-census}
plot.census <- function(x, cat_var, num_var, ...) {

  if (missing(cat_var) || missing(num_var)) {
    stop("plot.census() needs both a categorical variable (cat_var) and a numeric variable (num_var)")
  }
  if (!cat_var %in% names(x)) {
    stop(cat_var, " is not a column in this data")
  }
  if (!num_var %in% names(x)) {
    stop(num_var, " is not a column in this data")
  }

  ggplot(x, aes(x = get(cat_var), y = get(num_var), weight = PWGTP)) +
    geom_boxplot() +
    labs(x = cat_var, y = num_var, title = paste(num_var, "by", cat_var)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

To verify this function we used a small dataset with age and commute time as numeric, sex and educational attainment as categorical for North Carolina in 2024.

```{r build-example-data}
my_census_data <- getSingleYearData(
  year = 2024,
  numeric_vars = c("AGEP", "JWMNP"),
  categorical_vars = c("SEX", "SCHL"),
  geo_level = "State",
  geo_values = "37"
)
```

```{r summarize-example-data}
summary(my_census_data)
summary(my_census_data, numeric_vars = "AGEP", categorical_vars = "SEX")
```

```{r plot-example-data}
plot(my_census_data, cat_var = "SEX", num_var = "AGEP")
```

## Investigation: Household Language Across Geography

```{r investigation-setup}
# one representative state per region, standing in for that whole region
# so each pull stays the size of a single-state call instead of an entire
# multi-state region's worth of PUMS microdata
state_codes <- c("36", "17", "37", "06")  # New York, Illinois, North Carolina, California
region_names <- c("Northeast", "Midwest", "South", "West")

region_data <- lapply(state_codes, function(code) {
  getSingleYearData(
    year = 2024,
    numeric_vars = "AGEP",
    categorical_vars = "HHL",
    geo_level = "State",
    geo_values = code
  )
})
names(region_data) <- region_names
```

```{r region-summaries}
# summary() called once per region - four different census objects, same
# categorical_vars to get weighted household-language counts for each
region_summaries <- lapply(region_data, summary, categorical_vars = "HHL")
region_summaries
```

```{r hhl-proportions}
# convert each region's weighted counts to proportions so regions with
# different total populations are directly comparable. tapply() (used
# NOTE: inside summary.census()) returns NA, not 0, for a language level with
# nobody in it for that region, so treat those as 0 before summing, or
# sum() would return NA and wipe out every proportion.
hhl_proportions <- lapply(region_summaries, function(s) {
  counts <- s$categorical$HHL
  counts[is.na(counts)] <- 0
  round(counts / sum(counts), 3)
})
hhl_proportions
```

```{r non-english-share}
# 1: English
english_only_label <- loadVariableValues(2024)$HHL[["1"]]
english_only_label

non_english_share <- sapply(hhl_proportions, function(p) 1 - p[[english_only_label]])
non_english_share
```

```{r plot-south-hhl}
# plot() called with different inputs too: age spread by language group
# within a single region (South) 
plot(region_data$South, cat_var = "HHL", num_var = "AGEP")
```

```{r combine-all-regions}
all_regions <- bind_rows(lapply(region_names, function(nm) {
  one_region <- region_data[[nm]]
  class(one_region) <- setdiff(class(one_region), "census")
  one_region$REGION <- nm
  one_region
}))
all_regions$REGION <- factor(all_regions$REGION, levels = region_names)
class(all_regions) <- c("census", class(all_regions))
```

```{r plot-and-summarize-all-regions}
# vs. age spread by region overall, now that everything's combined
plot(all_regions, cat_var = "REGION", num_var = "AGEP")
summary(all_regions, categorical_vars = "HHL")
```